A fully labelled proof system for intuitionistic modal logics

### 2019-09-30 caleidoscope/caleidoscope.tex
soundness section

\begin{definition}
	Let $\M = \langle W, \rel, \le, V \rangle$ be a model and $\B, \Gamma \Rightarrow \Delta$ a sequent. We say that $\M \Vdash \B, \Gamma \Rightarrow \Delta$ if for all $\f$ such that:
	\begin{enumerate}
		\item For all $\labels{x}{A} \in \Gamma$, we have $\M, f(\lb x) \Vdash \fm A$.
		\item For all $\lb x, \lb y \in \B$, we have $f(\lb x) \rel f(\lb y)$.
		\item For all $\futs xy \in \B$, we have $f(\lb x) \le f(\lb y)$.
	\end{enumerate}
	then there exists $\labels{z}{B} \in \Delta$ such that $\M, f(\lb z) \Vdash \fm B$. We say that $\M \not \Vdash \B, \Gamma \Rightarrow \Delta$ if it is not the case that $\M \Vdash \B, \Gamma \Rightarrow \Delta$.
\end{definition}

@@

	\begin{center}
		\emph{For all model $\M$, if $\M \Vdash \B,\Left, \futs xy, \accs yz, \labels{x}{\BOX A}, \labels{z}{A} \Rightarrow \Right$, then $\M \Vdash \B, \Left, \futs xy, \accs yz, \labels{x}{\BOX A} \Rightarrow \Right$.}
	\end{center}

	By way of contradiction, we assume that there exists a model $\M_{0}$ such that $\M_{0} \not \Vdash \B, x \le y, yRz,\Left,\break \labels{x}{\BOX A} \Rightarrow \Right$, i.e $\M_{0} \Vdash \B$, $\M_{0} \Vdash \Left$, $\M_{0} \Vdash \futs xy$, $ \M_{0} \Vdash \accs yz$, $\M_{0} \Vdash \labels{x}{\BOX A}$ and  $\M_{0} \not \Vdash \Right$. Therefore, in particular from $\M_{0} \Vdash \labels{x}{\BOX A}$ we have that for all $y, z \in W$ such that $\futs xy$ and $\accs yz$, then $\M_{0}, f(\lb z) \Vdash \fm A$, i.e $\M_{0} \Vdash \labels{z}{\fm A}$.

\marianela{working on this}
	\marianela{
	Then assume there exists a model $\M_{0}$ such that $\M_{0} \not \Vdash \B, \Left, x \le y,\break yRz, \labels{x}{\BOX A} \Rightarrow \Right$, i.e $\M_{0} \Vdash \B$, $\M_{0} \Vdash \Left$, $\M_{0} \Vdash x \le y$, $ \M_{0} \Vdash yRz$, $\M_{0} \Vdash \labels{x}{\BOX A}$ and  $\M_{0} \not \Vdash \Right$. Therefore, in particular, from $\M_{0} \Vdash \labels{x}{\BOX A}$ we have that for all world $y, z$ such that $x \le y$ and $yRz$, then $\M_{0} \Vdash \labels{z}{A}$. Finally, we obtain that $\M_{0} \not \Vdash \B, \Left, x \le y, yRz, \labels{x}{\BOX A}, \labels{z}{A} \Rightarrow \Right$. Then, the rule $\llabrn\BOX$ is sound.

	\item Rule $\rlabrn\BOX$: By way of contradiction. Assume there exists a model $\M_{0}$ such that $\M_{0} \not \Vdash$ $conclusion$, i.e. $\M_{0} \not \Vdash \B, \Left \Rightarrow \Right, \labels{x}{\BOX A}$, which is the same as $\M_{0} \Vdash \vlan (\B,\Left)$ and  $\M_{0} \not \Vdash \vlor (\Right, \labels{x}{\BOX A})$. From $\M_{0} \not \Vdash \vlor (\Right, \labels{x}{\BOX A})$, we have that $\M_{0} \not \Vdash \Right$ and $\M_{0} \not \Vdash \labels{x}{\BOX A}$. Following the semantic definition of $\BOX$, there exists worlds $y, z$ in $\M_{0}$, such that $x \le y, yRz$ where $\M_{0} \not \Vdash \labels{z}{A}$. Therefore, $\M_{0} \not \Vdash premise$, i.e. $\M_{0} \not \Vdash \B, x \le y, yRz, \Left \Rightarrow \Right, \labels{z}{A}$. Then, the rule $\rlabrn\BOX$ is sound.

	\item Rule $\rlabrn\IMP$: By way of contradiction. Assume there exists a model $\M_{0}$ such that $\M_{0} \not \Vdash \B, \Left \Rightarrow \Right, \labels{x}{A \IMP B}$, i.e., in particular, $\M_{0} \not \Vdash \labels{x}{A \IMP B}$. Therefore, there exists $y$ such that $x \le y$ and $\M_{0} \Vdash \labels{y}{A}$ but $\M_{0} \not \Vdash \labels{y}{B}$. Then, we have $\M_{0} \not \Vdash \B, x \le y, \Left, \labels{y}{A} \Rightarrow \labels{y}{B}$. Finally, the rule $\rlabrn\IMP$ is sound.}
\end{itemize}

The other cases are similar.

@@

\section{Cut Admissibility}\label{sec:cut-elim}

\todo{I do not know how to organize this section. First cut elimination proof? (lemma before or after?) Then mon is admissible using cut proof?}

\lutz{some start below}

In this section we are going to prove the admissibility of cut for $\labIKp$.

@@

The \emph{height} of a derivation $\DD$, denoted by $\height\DD$, is \todo{finish}

We say that a rule is \emph{height-preserving admissible} if for every derivation $\DD$ of its premise there is a derivation $\DD'$ of its conclusion such that $\height{\DD'}\le\height\DD$. A rule is \emph{height-preserving invertible} if \todo{...}

@@

@@

\begin{lemma}
  The atomic version of $\llabrn{mon}$
  $$
  \vlinf{\labrn{mon_a}}{}{\lseq{\B, x \le x'}{\Left, \labels{x}{a}}\Right}{
    \lseq{\B, x \le x'}{\Left, \labels{x}{a}, \labels{x'}{a}}\Right}
  $$
  is height-preserving admissible for $\labIKp$.
\end{lemma}

@@

$\vlderivation{\vlhtr{\DD}{\lseq\B\Left\Right}}$

@@

\begin{theorem}\label{thm:cutfree-compl}
	%	Let $\CC$ be a set of geometric frame properties as in~\eqref{eq:cla-geometric} and $\labbrn{\CC}$ be the corresponding set of rules following schema~\eqref{eq:modal-grs}.
	%
	For any formula $A$, the following are equivalent.
	%
	\begin{enumerate}
		\item\label{i} $A$ is a theorem of $\IK$ 
		%
		\item\label{ii} $A$ is provable in $\labIKp +\labrn{cut}$ with %\quad
%		
%		\todo{}
		\smash{%\fbox{
			\scalebox{.9}{$\vliinf{\labrn{cut}}{}{\B_1, \B_2, \Left \SEQ \Right}{\B_1, \Left \SEQ \Right, \labels{z}{C}}{\B_2, \Left, \labels{z}{C} \SEQ \Right}$}
		}%}
		%

                \lutz{why is that multiplicative on $\B$ and additive on the rest???}
		\item\label{iii} $A$ is provable in $\labIKp$
		%
		\item\label{iv} $A$ is valid in every birelational frames %satisfying the properties in $\CC$.
	\end{enumerate}
\end{theorem}

>>>>

\begin{definition}
	Let $\M = \langle W, \rel, \le, V \rangle$ be a model and $\B, \Gamma \SEQ \Delta$ a sequent. We say that $\M \Vdash \B, \Gamma \SEQ \Delta$ if for all $\f$ such that:
	\begin{enumerate}
		\item For all $\labels{x}{A} \in \Gamma$, we have $\M, f(\lb x) \Vdash \fm A$. \marianela{should we say that this is equal to $\M \Vdash \labels{x}{A}$?}
		\item For all $\lb x, \lb y \in \B$, we have $f(\lb x) \rel f(\lb y)$.
		\item For all $\futs xy \in \B$, we have $f(\lb x) \le f(\lb y)$.
	\end{enumerate}
	then there exists $\labels{z}{B} \in \Delta$ such that $\M, f(\lb z) \Vdash \fm B$. We say that $\M \not \Vdash \B, \Gamma \SEQ \Delta$ if it is not the case that $\M \Vdash \B, \Gamma \SEQ \Delta$.
\end{definition}

@@

	\begin{center}
		\emph{For all model $\M$, if $\M \Vdash \B,\Left, \futs xy, \accs yz, \labels{x}{\BOX A}, \labels{z}{A} \SEQ \Right$, then $\M \Vdash \B, \Left, \futs xy, \accs yz, \labels{x}{\BOX A} \SEQ \Right$.}
	\end{center}

	By way of contradiction, we assume that there exists a model $\M_{0}$ such that $\M_{0} \not \Vdash \B, \futs xy, \accs yz,\Left,\break \labels{x}{\BOX A} \SEQ \Right$, i.e $\M_{0} \Vdash \B$, $\M_{0} \Vdash \Left$, $\M_{0} \Vdash \futs xy$, $ \M_{0} \Vdash \accs yz$, $\M_{0} \Vdash \labels{x}{\BOX A}$ and  $\M_{0} \not \Vdash \Right$. Therefore, in particular from $\M_{0} \Vdash \labels{x}{\BOX A}$ we have that for all $\lb y, \lb z \in W$ such that $\futs xy$ and $\accs yz$, $\M_{0}, f(\lb z) \Vdash \fm A$, i.e $\M_{0} \Vdash \labels{z}{\fm A}$. From this result, we can say that $\M_{0} \not \Vdash \B, \Left, \futs xy, \accs yz, \labels{x}{\BOX A}, \labels{z}{A} \SEQ \Right$. Thus, we have already proved that if there exists a model $\M_{0}$ such that $\M_{0} \not \Vdash \B, \futs xy, \accs yz,\Left,\break \labels{x}{\BOX A} \SEQ \Right$, then $\M_{0} \not \Vdash \B, \Left, \futs xy, \accs yz, \labels{x}{\BOX A}, \labels{z}{A} \SEQ \Right$. According to the definition, we proved that the rule $\llabrn\BOX$ is sound.

	\item Rule $\rlabrn\BOX$:

	In order to prove that the rule $\rlabrn\BOX$ of the system $\labIKp$ is sound, we need to show that:

	\begin{center}
		\emph{For all model $\M$, if $\M \Vdash \B, \futs xy, \accs yz, \Left \SEQ \Right, \labels{z}{A}$, then $\M \Vdash \B, \Left \SEQ \Right, \labels{x}{\BOX A}$.}
	\end{center}

	By way of contradiction, we want to prove that if there exists a model that does not satisfy the conclusion, then the same model does not satisfy the premise. Let us assume that there exists a model $\M_{0}$ such that $\M_{0} \not \Vdash \B, \Left \SEQ \Right, \labels{x}{\BOX A}$. In particular, $\M_{0} \not \Vdash \labels{x}{\BOX A}$, which means that there exist worlds $\lb y, \lb z$, such that $\futs xy$ and $\accs yz$ where $\M_{0} \not \Vdash \labels{z}{A}$. Finally, we have that $\M_{0} \Vdash \B, \futs xy, \accs yz, \Left$ and $\M_{0} \not \Vdash \labels{x}{\BOX A}, \labels{z}{A}$,  which is the same as $\M_{0} \not \Vdash \B, \futs xy, \accs yz, \Left \SEQ \Right, \labels{z}{A}$. Then, by way of contradiction, we proved that the rule $\rlabrn\BOX$ is sound.

	\item Rule $\rlabrn\IMP$:

	We display the proof of soundness for the rule $\rlabrn\IMP$ showing that:
	\begin{center}
		\emph{For all model $\M$, if $\M \Vdash \B, \futs xy, \Left, \labels{y}{A} \SEQ \Right \labels{y}{B}$, then $\M \Vdash \B, \Left \SEQ \Right, \labels{x}{A \IMP B}$.}
	\end{center}

	We prove this statement by way of contradiction. Assume there exists a model $\M_{0}$ such that $\M_{0}$ does not satisfy the conclusion, i.e. $\M_{0} \not \Vdash \B, \Left \SEQ \Right, \labels{x}{A \IMP B}$. In particular, $\M_{0} \not \Vdash \labels{x}{A \IMP B}$. Therefore, there exists $\lb y$ such that $\futs xy$ and $\M_{0}, f(\lb y) \Vdash A$ (i.e. $\M_{0} \Vdash \labels{y}{A}$) but $\M_{0}, f(\lb y) \Vdash B$, i.e. $\M_{0} \not \Vdash \labels{y}{B}$. Then, we have $\M_{0} \not \Vdash \B, \futs xy, \Left, \labels{y}{A} \SEQ \labels{y}{B}$. This means that the model $\M_{0}$ does not satisfy the premise and that is what we wanted to prove. Finally, we showed that the rule $\rlabrn\IMP$ is sound.

	\item Rule $\llabrn\DIA$:

	Following the same reasoning that we have been tabling, we define what means soundness of the rule $\llabrn\DIA$:

	\begin{center}
		\emph{For all model $\M$, if $\M \Vdash \B, \accs xy,  \Left, \labels{y}{A} \SEQ \Right$, then $\M \Vdash \B, \Left, \labels{x}{\DIA A} \SEQ \Right$.}
	\end{center}

	In order to prove this statement, we assume there exists a model $\M_{0}$ such that $\M_{0} \not \Vdash \B, \Left, \labels{x}{\DIA A} \SEQ \Right$ and we want to see that $\M_{0} \not \Vdash \B, \accs xy, \Left, \labels{y}{A} \SEQ \Right$.
	From our assumption, we have that from $\M_{0} \not \Vdash \B, \Left, \labels{x}{\DIA A} \SEQ \Right$ we obtain $\M_{0} \Vdash \B$, $\M_{0} \Vdash \Left$, $\M_{0} \Vdash \labels{x}{\DIA A}$ and $\M_{0} \not \Vdash \Right$. From $\M_{0} \Vdash \labels{x}{\DIA A}$, according to the definition of the $\DIA$ operator, we have that there exists a world $\lb y$ in $\M_{0}$ such that $\accs xy$ and $\M_{0} \Vdash \labels{y}{A}$. Therefore, we have that $\M_{0} \not \Vdash \B, \accs xy, \Left, \labels{y}{A} \SEQ \Right$ and we prove that the rule $\llabrn\DIA$ is sound.

	\item Rule $\rlabrn\DIA$:

	We want to prove that the rule $\rlabrn\DIA$ is sound, which means that:

	\begin{center}
		For all model $\M$, if $\M \Vdash \B, \accs xy, \Left \SEQ \Right, \labels{x}{\DIA A}, \labels{y}{A}$ then $\M \Vdash \B, \accs xy, \Left \SEQ \Right, \labels{x}{\DIA A}$
	\end{center}

	We prove this statement by way of contradiction. Let us assume there exists a model $\M_{0}$ such that $\M_{0} \not \Vdash \B, \accs xy, \Left \SEQ \Right, \labels{x}{\DIA A}$, we want to show that $\M_{0} \not \Vdash \B, \accs xy, \Left \SEQ \Right, \labels{x}{\DIA A}, \labels{y}{A}$.
	From our hypothesis we have that $\M_{0} \not \Vdash \labels{x}{\DIA A}$, i.e. for all world $\lb y \in \M_{0}$ where $\accs xy$ we have $\M_{0} \not \Vdash \labels{y}{A}$. Therefore, we have that $\M_{0} \not \Vdash \B, \accs xy, \Left \SEQ \Right, \labels{x}{\DIA A}, \labels{y}{A}$. Finally, assuming that $\M_{0}$ does not satisfy the conclusion, we proved that $\M_{0}$ does not satisfy the premise. Then we conclude that the rule $\rlabrn\DIA$ of the system $\labIKp$ is sound.
\end{itemize}

The other cases are similar and we leave them to the reader.

@@

\section{Cut Admissibility}\label{sec:cut-elim}
In this section we are going to prove the admissibility of cut for $\labIKp$.

@@

The \emph{height} of a derivation $\DD$, denoted by $\height\DD$, is the height of D when seen as a tree, i.e., the length of
the longest path in the tree from its root to one of its leaves.

We say that a rule is \emph{height-preserving admissible} if for every derivation $\DD$ of its premise there is a derivation $\DD'$ of its conclusion such that $\height{\DD'}\le\height\DD$. A rule is \emph{height-preserving invertible} if for every
derivation of the conclusion of the rule there are derivations for each of its premisses with at most the same height.

@@

\begin{proof}
  \todo{\begin{lemma}
  	If there exists a proof  $\vlderivation {\vlpd{\D}{}{\B, \Left \Rightarrow \Right}}$ then there exists a proof $\vlderivation {\vlpd {\Dw}{}{\B, xRy, u \le v, \Left, z \colon A \Rightarrow \Right, w \colon B}}$

  \end{lemma}

  \vspace{3mm}

  \begin{proof}
  	By induction on the height of $\D$.

  	For a proof of height 1:

  	if $\D$ = $\vlderivation {\vlin{\rn{id}}{}{\B, \Left, x \le y, x \colon a \Rightarrow \Right, y \colon a}{\vlhy {}}}$ then we take $\Dw$ to be $\vlderivation {\vlin{\rn{id}}{}{\B, \Left, x \le y, xRy, u \le v, x \colon a, z \colon A \Rightarrow \Right, w \colon B, y \colon a}{\vlhy {}}}$.
  	\vspace{2mm}

  	For a proof $\D$ of height greater than 1:

  	\begin{center}

  		$\vlderivation{\vlin {$r$}{}{\B, \Left \Rightarrow \Right}{\vlpd {\Done}{}{\B', \Left' \Rightarrow \Right'}}}$

  	Then by induction hypothesis there exists a proof

  	\begin{center}

  		$\vlderivation {\vlpd {\Dwone}{}{\B', xRy, u \le v, \Left', z \colon A \Rightarrow \Right', w \colon B}}$

  	Therefore, we have the proof

  	\begin{center}

  		$\Dw = \vlderivation {\vlin{}{}{\B, xRy, u \le v, \Left, z \colon A \Rightarrow \Right, w \colon B}{\vlpd {\Dwone}{}{\B', xRy, u \le v, \Left', z \colon A \Rightarrow \Right', w \colon B}}}$

  \end{proof}}
\end{proof}

@@

\begin{lemma}
  The atomic version of $\llabrn{mon}$
  $$
  \vlinf{\labrn{mon_a}}{}{\lseq{\B, \futs{x}{x'}}{\Left, \labels{x}{a}}\Right}{
    \lseq{\B, \futs{x}{x'}}{\Left, \labels{x}{a}, \labels{x'}{a}}\Right}
  $$
  is height-preserving admissible for $\labIKp$.
\end{lemma}

\begin{proof}
  	\begin{smallequation*}
  		\vlderivation{
  			\vlin{\labrn{mon_a}}{}{\B, \futs{x}{x'}, \futs{x'}{x''}, \Left, \labels{x}{a} \SEQ \Right, \labels{x''}{a}}{
  				\vlin{\labrn{id}}{}{\B, \futs{x}{x'}, \futs{x'}{x''}, \Left, \labels{x}{a}, \labels{x'}{a} \SEQ \Right, \labels{x''}{a}}{
  					\vlhy{}
  				}
  			}
  		}
  		\reducesto
  		\vlderivation{
  		\vlin{\color{red}\labrn{trans}}{}{\B, \futs{x}{x'}, \futs{x'}{x''}, \Left, \labels{x}{a} \SEQ \Right, \labels{x''}{a}}{\vlin{\labrn{id}}{}{\B, \futs{x}{x'}, \futs{x'}{x''}, \futs{x}{x''}, \Left, \labels{x}{a} \SEQ \Right, \labels{x''}{a}}{
  				\vlhy{}
  			}}
  	}
  	\end{smallequation*}
  	\todo{missing case}
%  	\begin{smallequation*}
%  		\vlderivation{
%  			\vlin{\labrn{mon_a}}{}{\B, \futs{x}{x'}, \Left, \labels{x}{a} \SEQ \Right, \labels{x'}{a}}{
%  				\vlin{\labrn{id}}{}{\B, \futs{x}{x'}, \labels{x}{a}, \labels{x'}{a} \SEQ \Right, \labels{x'}{a}}{
%  					\vlhy{}
%  				}
%  			}
%  		}
%  		\reducesto
%  		\vlinf{\labrn{id}}{}{\B, \futs{x}{x'}, \Left, \labels{x}{a} \SEQ \Right, \labels{x'}{a}}{}
%  		\qedhere
%  	\end{smallequation*}
\end{proof}

@@

%$\vlderivation{\vlhtr{\DD}{\lseq\B\Left\Right}}$

@@
